Notes of Basic Topolgy

### 2017-01-02 Basic Topology Notes/sections/ch9-DegreeAndLefschetzNumber.tex
[major] notes about Euler-Poincare lemma, Betti number, and free actions on spheres S^n when n is even.

%TODO ntoes for the next section.

>>>>

And yet another theorem left out by the teacher:
\begin{thm}
    If $n$ is even, then only $\Z_2$ and the trivial group can act freely on
    $S^n$.
\end{thm}
\begin{proof}
    Suppose a group $G$, which is not the trivial group, acts freely on $S^n$.
    Pick two elements $f,g$ of $G\setminus\{1\}$, they must have not fixed
    points, so they have $(-1)^{n+1}=-1$ degree. This means their composite
    $f\circ g$ have a degree of $(-1)\times(-1)=1$, whence we see that $f\circ
    g$ has fixed point. So $f\circ g= 1$. To repeat, $f\circ g=1$ for any
    $f,g\in G\setminus{1}$. So $G\cong \Z_2$.
\end{proof}
\begin{remark}
    Physically, this means that most reasonable symmetries, acting on $S^{2n}$ (such
    as a sphere $S^2$, the most common one in physics), are not free. They are
    subject to certain constraints such that there is always some point fixed
    when rotating the sphere.
\end{remark}
\subsection{Section 9.2 The Eular-Poincar\'e formula}
\label{sec:The-Eular-Poincare-formula}

Since we have skipped too many sections in the beginning, we lost a historical
background for this section. Therefore, I will be tacit in this section.

The first chapter of book\cite{book} introduced yet another topological invariant
to our list: the Euler characteristics. This section proves that the Eular
characteristics is actually a homotopy invariant.

The \nomen{Eular characteristics $\chi(K)$} of a simplicial complex is by
definition:
\begin{equation}
    \chi(K)=\sum_{q=0}^n (-1)^q \alpha_q
\end{equation}
where $n$ is the dimension of the simplicial complex $K$, $\alpha_q$ is the
number of $q$-simplexes in $K$. This sections defines the number $\beta_q$ as
the \textbf{$q$th Betti number of $K$}\nomenclature{$q$th Betti number of $K$,
$\beta_q$}{\nomrefpage}. It is the rank of free abelian part of $H_q(K)$
\footnote{
    recall that every finitely generated abelian group is isomorphic to a direct
    sum of primary cyclic groups (the torsion part) and infinite cyclic groups
    (the free part). A primary cyclic group is one whose order is a power of a
    prime. See more at
    \href{https://en.wikipedia.org/wiki/Finitely_generated_abelian_group}{wikipedia}.
}. If we allow, in the definition of homology groups, the coefficients being
rational, we can form $H_q(K,\Q)$, and we have $\beta_q$ is just the dimension
of $H_q(K,\Q)$ as a vector space over $\Q$ (lemma (9.9) in \cite{book}). Then
this section will shows:
\begin{thm}[Euler-Poincar\'e formula]
\begin{equation}
    \chi(K)=\sum_{q=0}^n (-1)^q \beta_q
\end{equation}
\end{thm}
and obviously
\begin{coro}
    Simplicial complexes whose polyhedra are homotopy equivalent have the same
    Euler characteristic.
\end{coro}
